Anomalous waves triggered by abrupt depth changes: laboratory experiments and truncated KdV statistical mechanics

### 2019-11-16 paper2_JNLS/paper.tex
NM

	More specifically, we assume continuity of the surface displacement, $\eta$, at the ADC \nick{(cite)}. Since the temporal fluctuations of $\eta$ match, waves crossing the ADC must rapidly adjust in wavelength so as to attain the same frequency in two different depths --- a well-understood phenomenon \nick{(cite)}. Of course, since the normalized domain $x \in [-\pi,\pi]$ is scaled on the characteristic wavelength, there is no change in the dimensionless wave-field $u(x,t)$, giving the condition
\begin{equation}
\label{detmatch}
u(x,t) \vert_{t=0^-} = u(x,t) \vert_{t=0^+}, 
%\quad \mbox{for } -\pi \le x \le \pi,
\qquad \mbox{\em Deterministic matching condition}
\end{equation}	
We call \eqref{detmatch} the deterministic matching condition to contrast with its statistical counterpart introduced below. This matching condition is employed in the deterministic simulations of the TKdV dynamical system \eqref{TKdV}. We note that in the alternate formulation mentioned earlier, it is the product $\depth^{1/4} \eta$ that would match at the ADC \cite{johnson1997modern}.

@@

	This statistical matching condition, originally derived by \maetal \cite{majda2019}, imposes a relationship between the two inverse temperatures $\thup$ and $\thdn$. In particular, we view $\thup$ as given by the random-state of the incoming wave field. Thus, while we will treat $\thup$ as a free parameter that can be varied to study various possible system states, the downstream $\thdn$ is determined directly by the matching condition \eqref{statmatch}, giving the functional dependence
\begin{equation}
\thdn = \transf \left( \thup \right)
\qquad \mbox{\em Transfer function}
\end{equation}
The {\em transfer function} $\transf$ will be a key link needed to relate the theory to experiments.

@@

For simply computing the transfer function, this weighted-sampling approach offers significant advantages over more sophisticated methods, such as such Markov Chain Monte Carlo (MCMC), in its ability to reuse the same samples of $\uhvec$ for several different values of $\thdn$. That is, we sample $\uhvec$ from the uniform distribution and compute the list of Hamiltonian values only once, then simply vary $\thdn$ in \eqref{rooteq} with the secant method until the root is found. An MCMC method, on the other hand, would require the sampling to restart from scratch for each value of $\thdn$, since the random steps in MCMC depend directly on the target distribution \eqref{Gibbs}. We will however, use an MCMC method to initialize direct numerical simulations of TKdV due to the superior efficiency for a specific, given value of $\theta$.

@@

Rearranging the TKdV system \eqref{TKdV} gives
\begin{align}
\label{TKdV2}
&\pd{\uL}{t} = -  \frac{1}{2} C_3 \drat^{-3/2} \, \pd{}{x} \Proj (\uL)^2 - C_2 \drat^{1/2} \, \frac{\partial^3 \uL}{\partial x^3} = \RHS[\uL]
%\qquad \text{for } x \in [-\pi,\pi]
\end{align}	
where we have represented the right-hand side by the operator $\RHS[\uL]$. We employ a pseudo-spectral discretization of \eqref{TKdV2}, with de-aliasing applied to the quadratic nonlinear term $\Proj (\uL)^2$ according to the standard 2/3-rule. That is, we first pad $\uhat_k$ with zeros for $\Lambda < \abs{k} \le 3\Lambda/2$, transform to physical space and square to obtain $\uL^2$ on a fine grid, then transform back to spectral space and truncate to obtain the Fourier coefficients of $\Proj (\uL)^2$
For simplicity, we denote these Fourier coefficients $\usqdah_k$,
\begin{equation}
\Proj (\uL)^2 = \sum_{\abs{k} \le \Lambda} \usqdah_k e^{i k x}
\end{equation}
With this discretization, \eqref{TKdV2} can be recast in spectral space as a nonlinear ODE system
\begin{equation}
\label{TKdVDisc}
\td{}{t} \uhat_k =  -\frac{1}{2} C_3 \drat^{-3/2} \, ik \usqdah_k + C_2 \drat^{1/2} \, ik^{3} \uhat_k = \RHSh_k
\end{equation}
The quadratic nonlinearity present in this system, $\usqdah_k$, mixes the modes during evolution. We note the third-order linear term may become stiff for large $\Lambda$.

@@

	For the initial conditions, $\uhvec^0$, of the direct numerical simulation we sample from the upstream Gibbs ensemble $\Gup$ in \eqref{Gibbs} with a prescribed inverse temperature $\thup$. We achieve this sampling via a Metropolis-Hasting Monte-Carlo algorithm as detailed in \cite{majda2019}.

>>>>

	More specifically, we assume continuity of the surface displacement, $\eta$, across the ADC \cite{whitham2011linear, rey1992propagation}. Since the propagation speed $c$ changes with depth, waves crossing the ADC must rapidly adjust in wavelength in order to match the oscillation frequency just upstream of the ADC \cite{whitham2011linear, rey1992propagation}. The normalized domain $x \in [-\pi,\pi]$ is scaled on the characteristic wavelength and so there is no change in the dimensionless wave-field $u(x,t)$, giving the condition
\begin{equation}
\label{detmatch}
u(x,t) \vert_{t=0^-} = u(x,t) \vert_{t=0^+}, 
%\quad \mbox{for } -\pi \le x \le \pi,
\qquad \mbox{\em Deterministic matching condition}
\end{equation}	
We call \eqref{detmatch} the deterministic matching condition to contrast with its statistical counterpart introduced below. This matching condition is employed in the deterministic simulations of TKdV \eqref{TKdV}. We note that in the alternate formulation mentioned earlier, it is the product $\depth^{1/4} \eta$ that would match at the ADC \cite{johnson1997modern}.

@@

	This statistical matching condition, originally derived by \maetal \cite{majda2019}, imposes a relationship between the two inverse temperatures $\thup$ and $\thdn$. In particular, we view $\thup$ as given by the random-state of the incoming wave field. Thus, while we will treat $\thup$ as a parameter that can be varied to study various possible system states, the downstream $\thdn$ is determined directly by the matching condition \eqref{statmatch}, giving the functional dependence
\begin{equation}
\thdn = \transf \left( \thup \right)
\qquad \mbox{\em Transfer function}
\end{equation}
The {\em transfer function} $\transf$ will be a key link needed to relate the theory to experiments.

@@

For simply computing the transfer function, this weighted-sampling approach offers significant advantages over more sophisticated methods, such as such Markov Chain Monte Carlo (MCMC), in its ability to reuse the same samples of $\uhvec$ for several different values of $\thdn$. That is, we sample $\uhvec$ from the uniform distribution and compute the list of Hamiltonian values only once, then simply vary $\thdn$ in \eqref{rooteq} with the secant method until the root is found. An MCMC method, on the other hand, would require the sampling to restart from scratch for each value of $\thdn$, since the random steps in MCMC depend directly on the target distribution \eqref{Gibbs}. We will however, use MCMC to initialize direct numerical simulations of TKdV due to the superior efficiency for a single, given value of $\theta$.

@@

Rearranging the TKdV system \eqref{TKdV} gives
\begin{align}
\label{TKdV2}
&\pd{\uL}{t} = -  \frac{1}{2} C_3 \drat^{-3/2} \, \pd{}{x} \Proj (\uL)^2 - C_2 \drat^{1/2} \, \frac{\partial^3 \uL}{\partial x^3} = \RHS[\uL]
%\qquad \text{for } x \in [-\pi,\pi]
\end{align}	
where we have represented the right-hand side by the operator $\RHS[\uL]$. We employ a pseudo-spectral discretization of \eqref{TKdV2}, with de-aliasing applied to the quadratic nonlinear term $\Proj (\uL)^2$ according to the standard 2/3-rule. That is, we first pad $\uhat_k$ with zeros for $\Lambda < \abs{k} \le 3\Lambda/2$, transform to physical space and square to obtain $\uL^2$ on a fine grid, then transform back to spectral space and truncate to obtain the Fourier coefficients of $\Proj (\uL)^2$
For simplicity, we denote these Fourier coefficients $\usqdah_k$,
\begin{equation}
\Proj (\uL)^2 = \sum_{\abs{k} \le \Lambda} \usqdah_k e^{i k x}
\end{equation}
With this discretization, \eqref{TKdV2} can be recast in spectral space as a nonlinear ODE system
\begin{equation}
\label{TKdVDisc}
\td{}{t} \uhat_k =  -\frac{1}{2} C_3 \drat^{-3/2} \, ik \usqdah_k + C_2 \drat^{1/2} \, ik^{3} \uhat_k = \RHSh_k
\end{equation}
The quadratic nonlinearity represented by $\usqdah_k$ mixes the modes during evolution. We note the third-order linear term may become stiff for large $\Lambda$.

@@

	For the initial conditions, $\uhvec^0$, of the direct numerical simulation we sample from the upstream Gibbs ensemble $\Gup$ \eqref{Gibbs} with a prescribed inverse temperature $\thup$. We achieve this sampling via a Metropolis-Hasting Monte-Carlo algorithm as detailed in \cite{majda2019}.
